Towards cross-middleware authentication and single sign-on for ARC Grid middleware

### 2009-02-01 doc/sec/papers/ISC09/ISC09.tex
More polishing

\title{Towards cross-middleware authentication and single sign-on for ARC Grid middleware
%\thanks{Grants or other notes
%about the article that should go on the front page should be
%placed here. General acknowledgments should be placed at the end of the article.}
}
%\subtitle{Do you have a subtitle?\\ If so, write it here}

%\titlerunning{Short form of title}        % if too long for running head

\author{Weizhong Qiang         \and
        Aleksandr Konstantinov %etc.
}

%\authorrunning{Short form of author list} % if too long for running head

\institute{F. Author \at
              first address \\
              Tel.: +123-45-678910\\
              Fax: +123-45-678910\\
              \email{fauthor@example.com}           %  \\
%             \emph{Present address:} of F. Author  %  if needed
           \and
           S. Author \at
              second address
}

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

\maketitle

\begin{abstract}
When pursuing the task of making access to Grids as simple as possible, security is one of the most important challenges in production Grid infrastructures, especially when the Grid applications span multiple administrative domains as well as heterogeneous Grid middlewares. A typical example is wide scale e-Science applications which need to coordinate resources shared among a number of independent institutions with different Grid middlewares deployed on these resources. In this paper, we describe security implementation and considerations used in the upcoming version of the Advanced Resource Connector (ARC) middleware, where the heterogeneity issue has been addressed. The main goal of ARC implementation in terms of security is to let the middleware be capable of interoperating with other Grid middlewares by leveraging on standard specifications. The key aspect of the work is to enhance the current proxy certificate based authentication and single sign-on by utilizing and enhancing the standardized Web Service specifications such as Security Assertion Markup Languages (SAML), single sign-on (SSO) profile and Web Services Security in order to achieve cross-middleware authentication and single sign-on.
\keywords{single sign-on \and delegation \and virtual organization \and ARC middleware }
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\section{Introduction} 
\label{sec:intro}
Grid provides the technology for wide-scale, cross-domain collaboration where users work with resources and share data which are distributed across administrative domains. This kind of collaboration is called Virtual Organization (VO)~\cite{IFoster01}. The resources that contribute to the VO can come from independent administrative domains, and even require different security approaches to access control due to usage of different underlying middlewares. The same heterogeneous situation applies to users who need to access the VO resources, because users holding one kind of a credential may need to utilize multiple resources with different security approaches. To solve the above issues, Grid middlewares should provide the following abilities:
The user which participates a VO should be able to access data and resources without being required to constantly do authentication at each resource or service;
The VO should be able to easily manage the privileges and memberships of the users;
The VO together with its resources should be able to easily manage the authorization, given the condition that the resources are owned by different administrative domains.

In the current Grid community, the Grid Security Infrastructure (GSI)~\cite{IFoster98} is the de facto standard for authentication and transport level security. It builds on the Public Key Infrastructure (PKI), an architecture based on the X.509 public key certificate. GSI implements some enhancement (such as certificate delegation) based on the standard SSL/TLS protocol. Mutual authentication is required by GSI, and is the default configuration for GSI based Grid deployment. The X.509 certificates are required for both the client and service sides in order to achieve mutual authentication. The X.509 certificates are issued by trusted third parties called certificate authorities (CA), and the CAs constitute trust federation which guarantees that two different X.509 certificates from different CAs can accomplish authentication against each other. Thus in the GSI based Grid system, if a user would like to access a Grid system, he/she should own a X.509 certificate which is issued by a CA that is trusted by other entities.

@@

Unfortunately, the existing Grid security solutions such as GSI and VOMS do not apply to Web Service based applications very well due to the following reasons:
The GSS API based confidential communication has not been adopted by Web Service implementations.
The Attribute Certificate has not been adopted by any Web Service implementation at all.
Web Service implementations commonly use the Web Services Security (WS-Security)~\cite{WSSeclink}, Security Assertion Markup Languages (SAML)~\cite{SAMLlink}, and standard SSL/TLS as well as standard X.509 certificate when considering security, which make them impossible to interoperate with the GSI based Grid middlewares.
Finally, identity delegation is a good solution for user or user’s representative (acting on behalf of the user) to move/travel from one resource to another. But since the identity delegation has been completely coupled with the GSI confidential communication process, it is not possible for the existing identity delegation solution to be used by Web Service applications.

@@

\section{Cross-Middleware Authentication and Single Sign-On}
\label{sec:siglesignon}
We investigated the authentication and single sign-on issue in the scenario when a user or user’s representative (acting on behalf of this user) needs to access resources across different middlewares. Our work focuses on the following five major aspects:
Integration of community authentication – to utilize the standard-compatible community authentication mechanism from the Shibboleth~\cite{Shiblink} Identity Provider software when users authenticate against the Grid services.
Short lived credentials service – to issue the short lived X.509 certificates for users authenticated using the community authentication, in order to interoperate with the services that depend on PKI based authentication.
Credential delegation service – to process the X.509 credential delegation as a normal Web Service, in order to provide single sign-on ability.
Message level authentication – to provide message level authentication by implementing WS-Security.
SAML token based delegation service – to process the delegation of an WS Security SAML token in order to provide single sign-on for SOAP message level authentication.

@@

\begin{figure}
\includegraphics[width=0.9\columnwidth]{SAML2SSO.png}
\caption{SAML2.0 SSO profile in ARC}
\label{fig:SAML2SSO}
\end{figure}
Here is the explanation of steps shown in Figure \ref{fig:SAML2SSO}:
1.The client uses the user agent interface to launch a HTTP request including the IdP name (to which the user belongs) to the service side. The endpoint of the SP (Service Provider) service is the same as that of the other target services, except the last part of the endpoint is “saml2sp” which is specific for pointing to the SP Service. Note that we use Identity Provider (IdP) name here to simplify the IdP discovery process in order to avoid the IdP discovery process, because we suppose that the user who would access the target services should better know where is he from initially.
2.The SP Service searches the metadata (we use the same metadata format as defined in the Shibboleth) and gets the location of the single sign-on service (hosted in IdP) and also the location of assertion consuming service (hosted in this SP itself). Then SP Service uses obtained information to compose and issue the SAML authentication request also by utilizing own X.509 certificate (note that in the SAML2.0 SSO profile, X.509 certificates are still needed for IdP and SP) and sends the response back to the user agent.
3.User agent sends the SAML authentication request to the Identity Provider.
4.Identity Provider requires an act of authentication. The authentication mechanism is outside of the SAML2.0 SSO profile. The Shibboleth IdP implementation uses login handlers for authentication. The current user agent implementation is compatible with the Username/Password login handler of Shibboleth IdP. Through the HTTP protocol, the user agent will feed IdP with the username/password which has been given by the caller of user agent interface.
5.Once the authentication has been succeeded, the IdP issues a SAML response including SAML assertion encrypted by destination SP’s public key, and then this SAML response will be delivered by the user agent to the Service Provider.
6.The SP Service verifies and checks the SAML response, decrypts and stores the SAML assertion into session/connection context. That assertion includes the AuthnStatement and AttributeStatement elements.
7.The WS client launches the Grid/Web Service request via the same connection as the one which is used by user agent to contact SP Service.
8.The Grid/Web Service checks the AuthnStatement stored in the session context to see if the session is still valid. If valid, service handles the  request and returns the response to the WS client. Note that in the current implementation the service requires WS client to reuse same TCP connection as the one used by user agent is step 5, in order to guarantee  validity of the SSO result.

The SP service and other functional service(s) are hosted by the same container, and they use the same X.509 credential for the transport layer security. The client authentication may be switched off, so that client doesn’t need to use any X.509 credential. Only the trusted CAs of SP and IdP services need to be configured on the client side to ensure SP and IdP can authentication.

@@

\begin{figure}
\includegraphics[width=0.9\columnwidth]{SAMLDelegationMessage.png}
\caption{SOAP message header with SAML delegation token inside}
\label{fig:SAMLDelegationMessage}
\end{figure}
The delegation process is similar to the process in the X.509 credential delegation service, with the following differences:
1.Transport level communication between delegation client and service is not based on mutual authentication between the proxy certificate and the service certificate; instead, it is based on mutual authentication between the service certificates.
2.Delegation service will not generate any certificate request; instead, the delegation client will pick the public key from the peer certificate after a successful SSL/TLS authentication. So step 2 of the delegation protocol is just some confirmation information which informs delegation client to proceed.
3.For those multiple intermediate services through which a SOAP message flows, delegation service does not need to appear, instead, it only needs to appear on the end of the SOAP message flow.
4.SAML Token might include one or a few <saml:Attribute> attributes which carry the attribute information related to the subject, instead of the Attribute Certificate which is used in the VOMS proxy certificate for carrying the attribute information.

Figure \ref{fig:SAMLDelegationMessage} shows the SAML Token (SAML Assertion) inside the SOAP message is changed due to delegation (we suppose X.509 token is used for attesting entity to authenticate against issuing authority). The SOAP message on the left side is the one which B sends to the relying party (intermediate service). B is the identity of the attesting entity’s X.509 token, while A is the identity of issuing authority, and the SOAP message is signed by B’s private key which is corresponding to the X.509 certificate that is used to create the X.509 token. On the right side, C is the identity of the delegation service (also C is the identity of the target service to which B will access), and now the SOAP message is signed by C’s private key which is corresponding to the target service’s certificate.

@@

\section{Related Work}
\label{sec:relatedwork}
Similar work that integrates Shibboleth and Grid middleware does exist. GridShib~\cite{VWelch05},\cite{TBarton06} uses Shibboleth to get entity’s attribute for the authorization of services developed in Globus Tookit. Unlike the work described in this paper, the GridShib does not use Shibboleth’s native login handlers for authentication to Shibboleth IdP, instead, a new login handler specific for X.509 based authentication is developed for the Shibboleth IdP, and a credential service is developed for acting as an online-CA, authenticating the user, and issuing short-lived X.509 credential. Moreover, this paper provides a single sign-on solution for the user with community credentials authenticating to Grid services by directly using SAML2.0 SSO profile.

SWITCH (The Swiss Education and Research Network) short-lived credential service (SLCS)~\cite{switchslcslink} provides a SLCS service and a client. The advantage of our work comparing to the SWTCH’s solution is that the SLCS service and client are based on Web Services, so that they can be implemented in any Web Service container while keeping interoperability.

In terms of delegation, apart from the credential delegation mechanism of GSI~\cite{IFoster98},~\cite{VWelch04}, several authors~\cite{MAhsant04} propose a delegation protocol based on the WS-Trust~\cite{WSTrustlink} in order to provide a standard and interoperable protocol for the delegation in Grids. WS-Trust will also be adopted by the ARC middleware to express the specifications required to define delegation protocol in a standard way. Gridsite project implements a X.509 credential delegation solution~\cite{GridSitelink} based on the Web Service, with which ARC delegation client can be made easily interoperable.

@@

\bibitem{IFoster01}
Foster I, Kesselman C, Tuecke S. The anatomy of the Grid: Enabling scalable virtual organizations, International Journal of Supercomputer Applications, 15(3), 200–222 (2001)

@@

\bibitem{AlfieriR05}
R. Alfieri, R. Cecchini, V. Ciaschini, L. dell’Agnello, A. Frohner, K. Lorentey, and F. Spataro. From gridmap-file to voms: managing authorization in a Grid environment. Future Generation Comp. Syst., 21(4), 549–558 (2005)

@@

\bibitem{MEllert07}
M. Ellert et al. Advanced resource connector middleware for lightweight computational grids, Future Generation computer systems, 23(2), 219–240 (2007)

@@

\bibitem{GAloisio05}
Giovanni Aloisio, Massimo Cafaro, Italo Epicoco, Daniele Lezzi, and Robert van Engelen, The GSI plug-in for gSOAP: Enhanced Security, Performance, and Reliability, ITCC conference 2005, IEEE Press, Volume I, pages 304-309.

>>>>

\title{Towards cross-middleware authentication and single sign-on for ARC Grid middleware
%\thanks{Grants or other notes
%about the article that should go on the front page should be
%placed here. General acknowledgments should be placed at the end of the article.}
}
%\subtitle{Do you have a subtitle?\\ If so, write it here}
\titlerunning{Authentication and single sign-on in ARC}        % if too long for running head
\author{Weizhong Qiang         \and
        Aleksandr Konstantinov %etc.
}

%\authorrunning{Short form of author list} % if too long for running head
\institute{W. Qiang \at
              Department of Physics \\
              University of Oslo, Norway\\
              \email{weizhong.qiang@fys.uio.no}           %  \\
%             \emph{Present address:} of F. Author  %  if needed
           \and
           A. Konstantinov \at
              \email{aleksandr.konstantinov@fys.uio.no}
}
\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor
\maketitle
\begin{abstract}
When pursuing the task of making access to Grids as simple as possible, security is one of the most important challenges in production Grid infrastructures, especially when the Grid applications span multiple administrative domains as well as heterogeneous Grid middlewares. A typical example is wide scale e-Science applications which need to coordinate resources shared among a number of independent institutions with different Grid middlewares deployed on these resources. In this paper, we describe security implementation and considerations used in the upcoming version of the Advanced Resource Connector (ARC) middleware, where the heterogeneity issue has been addressed. The main goal of ARC implementation in terms of security is to let the middleware be capable of interoperating with other Grid middlewares by leveraging on standard specifications. The key aspect of the work is to enhance the current proxy certificate based authentication and single sign-on by utilizing and enhancing the standardized Web Service specifications such as Security Assertion Markup Languages (SAML), single sign-on (SSO) profile and Web Services Security in order to achieve cross-middleware authentication and single sign-on.
\keywords{single sign-on \and delegation \and virtual organization \and ARC middleware }
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\section{Introduction} 
\label{sec:intro}
Grid provides the technology for wide-scale, cross-domain collaboration where users work with resources and share data which are distributed across administrative domains. This kind of collaboration is called Virtual Organization (VO)~\cite{IFoster01}. The resources that contribute to the VO can come from independent administrative domains, and even require different security approaches to access control due to usage of different underlying middlewares. The same heterogeneous situation applies to users who need to access the VO resources, because users holding one kind of a credential may need to utilize multiple resources with different security approaches. To solve the above issues, Grid middlewares should provide the following abilities:
\begin{itemize}
\item The user which participates a VO should be able to access data and resources without being required to constantly do authentication at each resource or service;
\item The VO should be able to easily manage the privileges and memberships of the users;
\item The VO together with its resources should be able to easily manage the authorization, given the condition that the resources are owned by different administrative domains.
\end{itemize}
In the current Grid community, the Grid Security Infrastructure (GSI)~\cite{IFoster98} is the de facto standard for authentication and transport level security. It builds on the Public Key Infrastructure (PKI), an architecture based on the X.509 public key certificate. GSI implements some enhancement (such as certificate delegation) based on the standard SSL/TLS protocol. Mutual authentication is required by GSI, and is the default configuration for GSI based Grid deployment. The X.509 certificates are required for both the client and service sides in order to achieve mutual authentication. The X.509 certificates are issued by trusted third parties called certificate authorities (CA), and the CAs constitute trust federation which guarantees that two different X.509 certificates from different CAs can accomplish authentication against each other. Thus in the GSI based Grid system, if a user would like to access a Grid system, he/she should own a X.509 certificate which is issued by a CA that is trusted by other entities.

@@

Unfortunately, the existing Grid security solutions such as GSI and VOMS do not apply to Web Service based applications very well due to the following reasons:
\begin{itemize}
\item The GSS API based confidential communication has not been adopted by Web Service implementations.
\item The Attribute Certificate has not been adopted by any aspects Web Service implementation at all.
\item Web Service implementations commonly use the Web Services Security (WS-Security)~\cite{WSSeclink}, Security Assertion Markup Languages (SAML)~\cite{SAMLlink}, and standard SSL/TLS as well as standard X.509 certificate when considering security, which make them impossible to interoperate with the GSI based Grid middlewares.
\item Finally, identity delegation is a good solution for user or user’s representative (acting on behalf of the user) to move/travel from one resource to another. But since the identity delegation has been completely coupled with the GSI confidential communication process, it is not possible for the existing identity delegation solution to be used by Web Service applications.
\end{itemize}

@@

\section{Cross-Middleware Authentication and Single Sign-On}
\label{sec:siglesignon}
We investigated the authentication and single sign-on issue in the scenario when a user or user’s representative (acting on behalf of this user) needs to access resources across different middlewares. Our work focuses on the following five major aspects:
\begin{itemize}
	\item Integration of community authentication -- to utilize the standard-compatible community authentication mechanism from the Shibboleth~\cite{Shiblink} Identity Provider software when users authenticate against the Grid services.
	\item Short lived credentials service – to issue the short lived X.509 certificates for users authenticated using the community authentication, in order to interoperate with the services that depend on PKI based authentication.
	\item Credential delegation service – to process the X.509 credential delegation as a normal Web Service, in order to provide single sign-on ability.
	\item Message level authentication – to provide message level authentication by implementing WS-Security.
	\item SAML token based delegation service – to process the delegation of an WS Security SAML token in order to provide single sign-on for SOAP message level authentication.
\end{itemize}

@@

\begin{figure}
\includegraphics[width=0.9\columnwidth]{SAML2SSO.png}
\caption{SAML2.0 SSO profile in ARC}
\label{fig:SAML2SSO}
\end{figure}
Here is the explanation of steps shown in Figure \ref{fig:SAML2SSO}:
\begin{enumerate}
\item The client uses the user agent interface to launch a HTTP request including the IdP name (to which the user belongs) to the service side. The endpoint of the SP (Service Provider) service is the same as that of the other target services, except the last part of the endpoint is “saml2sp” which is specific for pointing to the SP Service. Note that we use Identity Provider (IdP) name here to simplify the IdP discovery process in order to avoid the IdP discovery process, because we suppose that the user who would access the target services should better know where is he from initially.
\item The SP Service searches the metadata (we use the same metadata format as defined in the Shibboleth) and gets the location of the single sign-on service (hosted in IdP) and also the location of assertion consuming service (hosted in this SP itself). Then SP Service uses obtained information to compose and issue the SAML authentication request also by utilizing own X.509 certificate (note that in the SAML2.0 SSO profile, X.509 certificates are still needed for IdP and SP) and sends the response back to the user agent.
\item User agent sends the SAML authentication request to the Identity Provider.
\item Identity Provider requires an act of authentication. The authentication mechanism is outside of the SAML2.0 SSO profile. The Shibboleth IdP implementation uses login handlers for authentication. The current user agent implementation is compatible with the Username/Password login handler of Shibboleth IdP. Through the HTTP protocol, the user agent will feed IdP with the username/password which has been given by the caller of user agent interface.
\item Once the authentication has been succeeded, the IdP issues a SAML response including SAML assertion encrypted by destination SP’s public key, and then this SAML response will be delivered by the user agent to the Service Provider.
\item The SP Service verifies and checks the SAML response, decrypts and stores the SAML assertion into session/connection context. That assertion includes the AuthnStatement and AttributeStatement elements.
\item The WS client launches the Grid/Web Service request via the same connection as the one which is used by user agent to contact SP Service.
\item The Grid/Web Service checks the AuthnStatement stored in the session context to see if the session is still valid. If valid, service handles the  request and returns the response to the WS client. Note that in the current implementation the service requires WS client to reuse same TCP connection as the one used by user agent is step 5, in order to guarantee  validity of the SSO result.
\end{enumerate}
The SP service and other functional service(s) are hosted by the same container, and they use the same X.509 credential for the transport layer security. The client authentication may be switched off, so that client doesn’t need to use any X.509 credential. Only the trusted CAs of SP and IdP services need to be configured on the client side to ensure SP and IdP can authentication.

@@

\begin{figure}
\includegraphics[width=0.9\columnwidth]{SAMLDelegationMessage.png}
\caption{SOAP message header with SAML delegation token inside}
\label{fig:SAMLDelegationMessage}
\end{figure}
The delegation process is similar to the process in the X.509 credential delegation service, with the following differences:
\begin{enumerate}
	\item Transport level communication between delegation client and service is not based on mutual authentication between the proxy certificate and the service certificate; instead, it is based on mutual authentication between the service certificates.
	\item Delegation service will not generate any certificate request; instead, the delegation client will pick the public key from the peer certificate after a successful SSL/TLS authentication. So step 2 of the delegation protocol is just some confirmation information which informs delegation client to proceed.
	\item For those multiple intermediate services through which a SOAP message flows, delegation service does not need to appear, instead, it only needs to appear on the end of the SOAP message flow.
	\item SAML Token might include one or a few <saml:Attribute> attributes which carry the attribute information related to the subject, instead of the Attribute Certificate which is used in the VOMS proxy certificate for carrying the attribute information.
\end{enumerate}
Figure \ref{fig:SAMLDelegationMessage} shows the SAML Token (SAML Assertion) inside the SOAP message is changed due to delegation (we suppose X.509 token is used for attesting entity to authenticate against issuing authority). The SOAP message on the left side is the one which B sends to the relying party (intermediate service). B is the identity of the attesting entity’s X.509 token, while A is the identity of issuing authority, and the SOAP message is signed by B’s private key which is corresponding to the X.509 certificate that is used to create the X.509 token. On the right side, C is the identity of the delegation service (also C is the identity of the target service to which B will access), and now the SOAP message is signed by C’s private key which is corresponding to the target service’s certificate.

@@

\section{Related Work}
\label{sec:relatedwork}
Similar work that integrates Shibboleth and Grid middleware does exist. GridShib~\cite{VWelch05,TBarton06} uses Shibboleth to get entity's attribute for the authorization of services developed in Globus Tookit. Unlike the work described in this paper, the GridShib does not use Shibboleth's native login handlers for authentication to Shibboleth IdP, instead, a new login handler specific for X.509 based authentication is developed for the Shibboleth IdP, and a credential service is developed for acting as an online-CA, authenticating the user, and issuing short-lived X.509 credential. Moreover, this paper provides a single sign-on solution for the user with community credentials authenticating to Grid services by directly using SAML2.0 SSO profile.

SWITCH (The Swiss Education and Research Network) short-lived credential service (SLCS)~\cite{switchslcslink} provides a SLCS service and a client. The advantage of our work comparing to the SWTCH's solution is that the SLCS service and client are based on Web Services, so that they can be implemented in any Web Service container while keeping interoperability.

In terms of delegation, apart from the credential delegation mechanism of GSI~\cite{IFoster98,VWelch04}, several authors~\cite{MAhsant04} propose a delegation protocol based on the WS-Trust~\cite{WSTrustlink} in order to provide a standard and interoperable protocol for the delegation in Grids. WS-Trust will also be adopted by the ARC middleware to express the specifications required to define delegation protocol in a standard way. Gridsite project implements a X.509 credential delegation solution~\cite{GridSitelink} based on the Web Service, with which ARC delegation client can be made easily interoperable.

@@

\bibitem{IFoster01}
Foster I, Kesselman C, Tuecke S. The anatomy of the Grid: Enabling scalable virtual organizations, International Journal of Supercomputer Applications, 15(3), 200-222 (2001)

@@

\bibitem{AlfieriR05}
R. Alfieri, R. Cecchini, V. Ciaschini, L. dell’Agnello, A. Frohner, K. Lorentey, and F. Spataro. From gridmap-file to voms: managing authorization in a Grid environment. Future Generation Comp. Syst., 21(4), 549-558 (2005)

@@

\bibitem{MEllert07}
M. Ellert et al. Advanced resource connector middleware for lightweight computational grids, Future Generation computer systems, 23(2), 219-240 (2007)

@@

\bibitem{GAloisio05}
Giovanni Aloisio, Massimo Cafaro, Italo Epicoco, Daniele Lezzi, and Robert van Engelen, The GSI plug-in for gSOAP: Enhanced Security, Performance, and Reliability, ITCC conference 2005, IEEE Press, Volume I, pp. 304-309.
